---
title: "Class12"
author: "Jessica Gao (PID: 16939806)"
format: pdf
---

## Section 1 Proportion of G/G in a population

Downloaded a csv file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read the csv file
```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```
```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```
## Section 4. Population analysis

>Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table ("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```
```{r}
table(expr$geno)
```
```{r}
library(ggplot2)
```

Let's make a boxplot

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 

```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno)+
  geom_boxplot(notch=TRUE)
```
A|A seems to be related to an increase in the expression value, and G|G is related to a decrease of the expression value. Yes the SNP effect the expression of ORMDL3. 


