---
title: "Class18: Pertussis Mini Project"
author: "Jessica Gao (PID:A16939806)"
format: pdf
---

Pertussis, also known as the whopping cough is a deadly lung infection caused by the bacteria B. Pertussis

The CDC tracks Pertussis causes around the U.S. 
https://tinyurl.com/pertussiscdc 

We can "scrape" this data using the R **datapasta** package.

>Q1 With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r, include=FALSE}
cdc <- data.frame(
  year = c(1922L,1923L,1924L,1925L,
           1926L,1927L,1928L,1929L,1930L,1931L,
           1932L,1933L,1934L,1935L,1936L,
           1937L,1938L,1939L,1940L,1941L,1942L,
           1943L,1944L,1945L,1946L,1947L,
           1948L,1949L,1950L,1951L,1952L,
           1953L,1954L,1955L,1956L,1957L,1958L,
           1959L,1960L,1961L,1962L,1963L,
           1964L,1965L,1966L,1967L,1968L,1969L,
           1970L,1971L,1972L,1973L,1974L,
           1975L,1976L,1977L,1978L,1979L,1980L,
           1981L,1982L,1983L,1984L,1985L,
           1986L,1987L,1988L,1989L,1990L,
           1991L,1992L,1993L,1994L,1995L,1996L,
           1997L,1998L,1999L,2000L,2001L,
           2002L,2003L,2004L,2005L,2006L,2007L,
           2008L,2009L,2010L,2011L,2012L,
           2013L,2014L,2015L,2016L,2017L,2018L,
           2019L,2020L,2021L,2022L, 2024L),
  cases = c(107473,164191,165418,152003,
                                   202210,181411,161799,197371,
                                   166914,172559,215343,179135,265269,
                                   180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,
                                   120718,68687,45030,37129,60886,
                                   62786,31732,28295,32148,40005,
                                   14809,11468,17749,17135,13005,6799,
                                   7717,9718,4810,3285,4249,3036,
                                   3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,
                                   3589,4195,2823,3450,4157,4570,
                                   2719,4083,6586,4617,5137,7796,6564,
                                   7405,7298,7867,7580,9771,11647,
                                   25827,25616,15632,10454,13278,
                                   16858,27550,18719,48277,28639,32971,
                                   20762,17972,18975,15609,18617,
                                   6124,2116,3044,35493)
)
```


```{r}
head(cdc)
```

>Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)

ggplot(cdc)+
  aes(year,cases)+
  geom_line()
```


>Q2 Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
library(ggplot2)

ggplot(cdc)+
  aes(year,cases)+
  geom_line()+
  geom_vline(xintercept=1946, col="blue")+
  geom_vline(xintercept=1996, col="red")+
  geom_vline(xintercept=2020, col="gray")+
  geom_vline(xintercept=2003, col="darkgreen")
```

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

There were high cases numbers before the first wP(whole-cell) vaccine roll out in 1946 then a rapid decline in case numbers until 2004 when we have our first large-scale outbreaks of pertussis again. There is also a notable COVID-19 related dip and recent rapid rise. 

So the question is what is different about the immune response to the infection if you had a older version wP vaccine versus the newer aP vaccine.  

##CMI-PB (Computational Models of Immunity Pertussis Boost)

The CMI-PB project aims to address this key question: what is the different between aP and wP individuals. 

We can get all the data from this ongoing project via JSON API calls.
For this we will use the **jsonlite** package
```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject",
                     simplifyVector=TRUE)
head(subject)

```

>Q. How many individuals "subjects" are in this dataset?

```{r}
nrow(subject)
```

>Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```
## Working with dates

```{r}
library(lubridate)
```

>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
subject$age <- today() - ymd(subject$year_of_birth)
```

```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
# wP
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

>Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

>Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

```{r}
# Or use wilcox.test() 
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```


Obtain more data from CMI-PB

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen",
                      simplifyVector = TRUE)
ab_data <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer",
                     simplifyVector = TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_data)
```

I now have three tables of data from CMI-PB: 'subject,' 'specimen,' and 'ab_data.' I need to join these tables so I will have all tge info I need to work with.

For this we will use the 'inner_joint()' function from the **dplyr** packages.

>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)

meta <- inner_join(subject,specimen)
head(meta)
```

```{r}
dim(subject)
dim(specimen)
```

```{r}
dim(meta)
```
```{r}
head(meta)
```
>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

Now we can join our 'ab_data' table to 'meta' so we have all the info we need about antibody levels. 

```{r}
abdata <- inner_join(meta, ab_data)
head(abdata)
```

>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?
>Q. How many different antibody isotypes are there in this dataset?

```{r}
length(abdata$isotype)
```

```{r}
table(abdata$isotype)
```

>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

```{r}
table(abdata$antigen)
```

I want a plot of antigen levels across the whole dataset.

>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(abdata)+
  aes(MFI, antigen)+
  geom_boxplot()
```
```{r}
ggplot(abdata)+
  aes(MFI_normalised, antigen)+
  geom_boxplot()
```

>Q. Are there differences at this whole-dataset level between aP and wP?

Antigens like FIM2/3, PT, FELD1 have quite a large range of values. Others like Measels don't show much activity.
These ones that has large range values are in the wP vaccines.

```{r}
ggplot(abdata)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot()
```

```{r}
ggplot(abdata)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot()+
  facet_wrap(~infancy_vac)
```

## Examine IgG Ab titer levels

For this I need to select out just isotype IgG.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot()
```


```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot()+
  facet_wrap(~infancy_vac)
```

>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

PT, FELD1, and FIM2/3 show differences in the level of IgG antibody titers, their MFI_normalised is much higher than the other antigens. This is because they are the antigens present within the vaccines to give immunity to the patient's body while the other antigens were used more as control groups and not within the vaccines. 

Digging in further to look at the time course of IgG Isotype PT antigen levels across aP and wP individuals:

```{r}
##Filter to include 2021 data only
abdata.21 <- abdata |>  filter(dataset == "2021_dataset")
##Filter to look at IgG PT data only
pt.igg <- abdata.21 |>
  filter(isotype == "IgG",  antigen == "PT")
##Plotting and color by infancy_vac(wP vs aP)
  ggplot(pt.igg) +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```





