---
title: "Comparative structure analysis"
author: "Jessica Gao (PID:16939806)"
format: html
---

"install.packages("bio3d")
install.packages("devtools")
install.packages("BiocManager")

BiocManager::install("msa")
devtools::install_bitbucket("Grantlab/bio3d-view")"

>Q10. Which of the packages above is found only on BioConductor and not CRAN?

"msa"

>Q11. Which of the above packages is not found on BioConductor or CRAN?:

"Grantlab/bio3d-view"

>Q12. True or False? Functions from the devtools package can be used to install packages from GitHub and BitBucket?

True

Here we run through a complete "pipeline" of structure analysis that begins with a single sequence identifier and ends in a PCA analysis. 

```{r}
library(bio3d)
```

##Step 1. Retrieve sequence
Retrieve a sequence for the protein we are interested in. We will take ADK "1ake_A"

```{r}
id <- "1ake_A"
aa <- get.seq(id)
aa
```

>Q13. How many amino acids are in this sequence, i.e. how long is this sequence?

214

##Step 2. Blast search
Running a BLAST search of the PDB for all related sequences to our input 'aa'

```{r}
blast <- blast.pdb(aa)
```

```{r}
hits <- plot(blast)
```
##Step 3. Download
These are our "top hits" i.e. all the sturctures in the PDB database related to our input sequence. 

```{r}
hits$pdb.id
```

```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

Bioconductor has packages for bioinformatics 
 
## Step 4. Align and superpose

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```
Delete this chunk? too long
```{r}
pdbs
```
 
## Step5. PCA

Use PCA to make sense of these complicated structure relationships. 

```{r}
pc <- pca(pdbs)
plot(pc)
```
There are three main clusters. 

Make a trajectory (movie) of the main confrontational changes captured by PC1. We will use the 'mktrj()' function for this...

```{r}
mktrj(pc, file="pca_result.pdb")
```

Back of the envelope comparasion of the PDB size to UniProt

```{r}
uniprot <- 253206171
pdb <- 231029
pdb/uniprot*100
```


 